### 2020-03-20 CLT_Preliminaries/CLT.tex
update delta=eps power

\input{ex_shared}

\begin{document}
\maketitle

\textbf{Abstract.} We study the problem of drift estimation for two-scale continuous time series. We set ourselves in the framework of overdamped Langevin equations, for which a single-scale surrogate homogenized equation exists. In this setting, estimating the drift coefficient of the homogenized equation requires pre-processing of the data, often in the form of subsampling. We avoid subsampling by filtering the data with an appropriate kernel function and compute maximum likelihood estimators based on the filtered process. We show that the estimators we propose are asymptotically unbiased and demonstrate numerically the advantages of our method with respect to subsampling. Finally, we show how our filtering methodology can be combined with Bayesian techniques and provide a full uncertainty quantification of the inference procedure.

\textbf{AMS subject classifications.} 62F15, 65C30, 62M05, 74Q10.

\textbf{Keywords.} Parameter inference, diffusion process, data-driven homogenization, filtering, Bayesian inference, Langevin equation.

\section{Introduction}

Efficient parameter estimation for stochastic models is essential in a wide range of applications in natural and social sciences. In several areas, the data originate from phenomena which vary continuously in time and which are endowed with a multiscale structure. This is the case, for example, in molecular dynamics, oceanography and atmosphere science or in econometrics. Frequently, it is desirable in this areas to infer from data a simpler model which captures effectively large-scale structures, or slow variations, disregarding small-scale fluctuations or treating them as a source of noise. The mismatch between the data and their desired slow-scale representation is a typical instance of a problem of model misspecification, which, if ignored or mistreated, can lead to wrong solutions. Indeed, the data, coming from the full dynamics, are compatible with the coarse-grained model only at the time scales at which the effective dynamics is valid.

In this paper we consider a simple multiscale setting arising from models of molecular dynamics, with a complete separation between a fast and a slow scale. In particular, we consider diffusion processes for which the confining potential has slow variations, and whose motion is perturbed by a fast-scale potential with rapid periodic and bounded oscillations. Given this simple class of model problems, we are interested in determining the drift coefficient of an equation of the overdamped Langevin type in which the fast-scale potential is eliminated. The theory of homogenization guarantees that such a single-scale equation can be uniquely determined, and our goal is therefore to obtain effective coarse-grained dynamics from data consistently with respect to the homogenization result.

Several methods to take into account model misspecification in multiscale frameworks as above exist. For diffusion processes, the proposed approaches rely in different measures to subsampling, which has proved itself to some extent effective in many applications, but which requires nevertheless precise knowledge of how separated the two characteristic time scales are. Robustness of this methodology is dubious, too, as inference results tend to be extremely sensitive to the subsampling rate.

In the rest of the introduction, we first give a brief overview of the existing literature on the topic of deterministic and stochastic multiscale inference problems, then introduce our novel methodology and its favourable properties and conclude with an outline of this paper.

\subsection{Literature Review}

For simple models in molecular dynamics, the effect of model misspecification was studied in a series of papers \cite{ABT10,ABJ13,PaS07,PPS09,PPS12,GaS17,GaS18} under the assumption of scale separation. In particular, for Brownian particles moving in two-scale potentials it was shown that, when fitting data from the full dynamics to the homogenized equation, the maximum likelihood estimator (MLE) is asymptotically biased \cite[Theorem 3.4]{PaS07}. To be more precise, in the large sample size limit the MLE converges to the coefficients of the unhomogenized equation, rather than to those of the homogenized one. The bias of the MLE can be eliminated by subsampling at an appropriate rate, which lies between the two characteristic time scales of the problem \cite[Theorems 3.5 and 3.6]{PaS07}.

Similar techniques can be employed in econometrics, in particular for the estimation of the integrated stochastic volatility in the presence of market micro-structure noise. In this case, too, the data have to be subsampled at an appropriate rate \cite{AMZ05,OSP10}. The correct subsampling rate can be in some instances rather extreme with respect to the frequency of the data and lead to get rid of more than $99\%$ of the data. As the intuition suggests, this increases significantly the bias of the estimator, which is usually taken care of with additional bias corrections and variance reduction procedures. The need of such methodology is accentuated by data being obtained at high-frequency \cite{AiJ14,ZMA05}.

The problem of extracting large-scale variations from multiscale data is studied in atmosphere and ocean science. In this field, too, subsampling the data is necessary to obtain an accurate coarse-grained model \cite{CoP09,YMV19}.

The necessity to subsample the data can be alleviated by using appropriate martingale estimators, as was done in \cite{KPK13,KKP15}. This class of estimators can be applied to the case where the noise is multiplicative and also given by a deterministic chaotic system, as opposed to white noise. Estimators of this family have been applied to time series from paleoclimatic data and marine biology and augmented with appropriate model selection methodologies \cite{KPP15}.

Inference of diffusion processes can be naturally performed under a Bayesian perspective. If one focuses on the drift coefficient, the form of the likelihood function guarantees, under a Gaussian prior hypothesis, that the posterior distribution is itself a Gaussian. The versatility of the Bayesian approach in the infinite-dimensional case \cite{Stu10, DaS16} gives the possibility to extend the problem of inferring the drift of a diffusion process to the non-parametric case \cite{PSV09, PSZ13}.

The issue of model misspecification in inverse problems with a multiscale structure has been treated in the context of partial differential equations, too. In particular, it has been shown that it is possible to infer a coarse-grained equation from data coming from the full model and to retrieve asymptotically the correct result \cite{NPS12}. A series of papers \cite{AbD18, AbD19, AGZ19} focuses on retrieving the full model when the multiscale coefficient is endowed with a specific parametrized structure. Being these problems ill-posed, the latter is achieved via Tikhonov regularization \cite{AbD19,NPS12}, adopting a Bayesian approach \cite{AbD18, NPS12} or exploiting techniques of Kalman filtering \cite{AGZ19}. In \cite{AbD18,AGZ19}, the authors highlight the need of accounting explicitly for the modelling error due to homogenization and apply statistical techniques taken from \cite{CDS18,CES14}.

\subsection{Our contributions}
In this paper, we bypass the issue of subsampling by implementing an appropriate filtering methodology. In particular, we show that smoothing the data coming from the multiscale model with an appropriate linear time-invariant filter of the exponential family allows to retrieve the drift coefficient of the homogenized equation. The methodology we present is not involved computationally, easy to implement in practice and presents two main advantages:
\begin{enumerate}
	\item the filter's kernel depends on two parameters which can be tuned to obtain more robust results and which can be interpreted as analogous to the subsampling rate. Nevertheless, while the MLE is extremely sensitive with respect to the latter, our filtering methodology is robust with respect to its parameters, and can therefore be applied as a black-box tool for parameter estimation,
	\item the entire stream of data is employed, which, in practice, enhances the quality of the filter-based MLE in terms of bias. Moreover, avoiding subsampling and thus discretising the data allows us to employ continuous-time theoretical tools.
\end{enumerate}
After having deduced unbiased estimators, we focus on how to insert the filtering methodology into a Bayesian approach. In particular, we analyse the effects of modifying the likelihood function if the smoothed trajectory is inserted in an appropriate manner. Under mild hypotheses on the filter parameters, we are able to show that the posterior distributions obtained with our methodology are asymptotically consistent with respect to the drift parameter of the homogenized equation. To our knowledge, this is the first instance of a successful Bayesian approach applied to multiscale diffusion processes.

\subsection{Outline}
The rest of the paper is organised as follows. In Section \ref{sec:Setting} we introduce the problem and lay the basis of our analysis setting the main assumptions and notation. In Section \ref{sec:Filter} we present our filtering methodology, with a particular focus on ergodic properties, on multiscale convergence and, naturally, on the properties of our estimators. In Section \ref{sec:Bayesian} we introduce the Bayesian framework and show how it can be enhanced with filtering techniques. Finally, in Section \ref{sec:NumExp} we demonstrate the effectiveness of our methodology via a series of numerical experiments.

\section{Problem setting}\label{sec:Setting}

In this section, we introduce the class of diffusion processes which we treat in this paper and the classical methodology employed for the estimation of the drift. Let $\epl > 0$ and let us consider the one-dimensional multiscale stochastic differential equation (SDE)
\begin{equation}\label{eq:SDE_MS}
	\d X_t^\epl = -\alpha \cdot V'(X_t^\epl) \dd t - \frac1\epl p'\left(\frac{X_t^\epl}\epl\right) \dd t + \sqrt{2\sigma} \dd W_t,
where, given a positive integer $N$, we have that $\alpha \in \R^N$ and $\sigma > 0$ are the drift and diffusion coefficients respectively and $W_t$ is a standard one-dimensional Brownian motion. The functions $V\colon \R \to \R^N$ and $p\colon \R \to \R$ correspond to the slow-scale and the fast-scale confining potentials. In particular, we assume 
\begin{equation}\label{eq:Potential}
	V(x) = \begin{pmatrix} V_1(x) & V_2(x) & \cdots & V_N(x) \end{pmatrix}^\top,
for smooth functions $V_i\colon \R \to \R$, $i = 1, \ldots, N$. Moreover, we assume $p$ to be smooth and periodic of period $L$. Theory of homogenization \cite[Chapter 3]{BLP78} guarantees the existence of an SDE of the form
\begin{equation}\label{eq:SDE_HOM}
	\d X_t = - A \cdot V'(X_t) \dd t + \sqrt{2\Sigma} \dd W_t,
where $W_t$ is the same Brownian motion as in \eqref{eq:SDE_MS}, such that $X_t^\epl \to X_t$ for $\epl\to 0$ in law as random variables in $\mathcal C^0([0, T]; \R)$. In particular, we have $A = K\alpha$ and $\Sigma = K \sigma$, where the coefficient \corr{$0<K<1$} is given by the formula
\begin{equation}\label{eq:K_HOM}
	K = \int_0^L (1 + \Phi'(y))^2 \, \mu(\d y),
with 
	\mu(\d y) = \frac1Z e^{-p(y)/\sigma} \dd y, \quad\text{where}\quad Z = \int_0^L e^{-p(y)/\sigma} \dd y,
and where the function $\Phi$ is the unique solution with zero-mean with respect to the measure $\mu$ of the elliptic partial differential equation
\begin{equation}\label{eq:CellProblem}
	-p'(y)\Phi'(y) + \sigma \Phi''(y) = p'(y), \quad 0 \leq y \leq L,
endowed with periodic boundary conditions. \corr{Let us remark that in this one-dimensional setting it is possible to determine $\Phi$ explicitly, and the homogenization coefficient $K$ is given by
	K = \frac{L^2}{Z\widehat Z},
	Z = \int_0^L e^{-p(y)/\sigma} \dd y, \quad \widehat Z = \int_0^L e^{p(y)/\sigma} \dd y.
\end{equation}}
We now briefly present the classical methodology for estimating the drift coefficient. Let $T > 0$ and let $X^\epl \defeq (X^\epl_t, 0\leq t \leq T)$ be a realization of the solution of \eqref{eq:SDE_MS} up to final time. Girsanov's change of measure formula applied to \eqref{eq:SDE_HOM} allows to write the likelihood of $X^\epl$ given a drift coefficient $A$ as
\begin{equation}\label{eq:Likelihood}
p(X^\epl \mid A) = \exp\left(-\frac{I(X^\epl\mid A)}{2\Sigma} \right), 
I(X^\epl \mid A) = \int_0^T A \cdot V'(X^\epl_t) \dd X^\epl_t + \frac12 \int_0^T \left( A \cdot V'(X^\epl_t) \right)^2 \dd t.
Minimizing the functional $I(X^\epl \mid A)$ with respect to $A$ therefore gives the maximum likelihood estimator (MLE) of $A$, which can be formally computed in closed form as
\begin{equation}\label{eq:MLE}
	\widehat A^\epl(T) \defeq \arg \min_{A \in \R^N} I(X^\epl \mid A) = M^{-1}h,
where $M\in\R^{N\times N}$ and $h\in\R^N$ are defined as
\begin{equation}\label{eq:MandH}
M = \frac1T \int_0^T V'(X^\epl_t) \otimes V'(X^\epl_t) \dd t, \quad h = \frac1T \int_0^T V'(X^\epl_t) \dd X^\epl_t,
and where $\otimes$ denotes the outer product in $\R^N$. Let us now state the assumptions which will be employed throughout the rest of our work. In particular, we consider the same dissipative setting as \cite[Assumption 3.1]{PaS07}.
\begin{assumption}\label{as:regularity} The potentials $p$ and $V$ satisfy
	\begin{enumerate}
		\item $p \in \mathcal C^\infty(\R) \cap L^\infty(\R)$ and is $L$-periodic for some $L > 0$,
		\item\label{as:regularity_diss} $V_i \in \mathcal C^\infty(\R)$ for all $i=1, \ldots, N$ is polynomially bounded from above and bounded from below,  and there exist $a,b > 0$ such that
		-\alpha \cdot V'(x) x \leq a - bx^2.
		\item $V'$ is Lipschitz continuous, i.e. there exists a constant $C > 0$ such that
			\norm{V'(x) - V'(y)}_2 \leq C\abs{x - y},
		\item for all $T > 0$, the symmetric matrix $M$ is positive definite and there exists $\bar \lambda > 0$ such that $\lambda_{\min}(M) \geq \bar \lambda$.
	\end{enumerate}
\end{assumption}

\corr{
\begin{remark}\label{rem:regularity_diss} In the following, in particular in the proof of Lemma \ref{lem:ergodicity}, we will employ Assumption \ref{as:regularity}\ref{as:regularity_diss} for the whole drift of the SDE \eqref{eq:SDE_MS}, i.e., the function 
	\begin{equation}
		V^\epl(x) \defeq \alpha \cdot V(x) - p\left(\frac{x}{\epl}\right).
	\end{equation}
	Since $p \in L^\infty$, the assumption above is sufficient for $V^\epl$ to satisfy Assumption \ref{as:regularity}\ref{as:regularity_diss} with different values for $a$ and $b$. In particular, if the condition holds for $\alpha \cdot V$ for a given $b > 0$, then one can choose a new value for the same coefficient which is arbitrarily close to the original one for $V^\epl$.
\end{remark}
}

Under these assumptions, the MLE given in \eqref{eq:MLE} is indeed the unique minimizer of the likelihood function, as shown in \cite[Theorem 2.4]{PSV09}.

Given the convergence of $X^\epl_t \to X_t$ in the space of continuous stochastic processes, one would expect that the MLE \eqref{eq:MLE} would be asymptotically unbiased for the drift coefficient $A$ of the homogenized equation \eqref{eq:SDE_HOM}. Instead, it is possible to prove that in the asymptotic limit for $T \to \infty$ and $\epl \to 0$, the MLE tends to the drift coefficient $\alpha$ of the unhomogenized equation \eqref{eq:SDE_MS}. We report here this result, whose proof can be found for the case $N = 1$ in \cite[Theorem 3.4]{PaS07}. Let us remark that the proof for $N > 1$ follows directly from the one-dimensional case.

\begin{theorem}\label{thm:Bias} Let Assumption \ref{as:regularity} hold and let $X^\epl_0$ be distributed according to the invariant measure of the process $X^\epl$ solution of \eqref{eq:SDE_MS}. Then
	\begin{equation}
		\lim_{\epl \to 0}\lim_{T \to \infty} \widehat A^\epl(T) = \alpha, \quad \text{a.s.},
	\end{equation}
	where $\alpha$ is the drift coefficient of equation \eqref{eq:SDE_MS}.
\end{theorem}

As anticipated in the introduction, the main tool for obtaining unbiased estimators in the literature is subsampling the data. In particular, let the dimension of the parameter $N = 1$, let $\delta > 0$ and let $T = n\delta$ with $n$ a positive integer. Then, a subsampled estimator for $A$ is given by
\begin{equation}
	\widehat A^\epl_\delta(T) = - \frac{\sum_{j=0}^{n-1} V'(X^\epl_{j\delta})\left(X^\epl_{(j+1)\delta} - X^\epl_{j\delta}\right)}{\delta \sum_{j=0}^{n-1} V'(X^\epl_{j\delta})^2},
\end{equation}
which is a discretized version of $\widehat A^\epl(T)$. It is possible to show \cite[Theorem 3.5]{PaS07} that choosing $\delta = \epl^\zeta$ with $\zeta \in (0, 1)$ and if $T$ is sufficiently big, then $\widehat A^\epl_\delta(T)$ is an asymptotically unbiased estimator of $A$ with respect to $T \to \infty$ and $\epl \to 0$, in probability. Despite being widely employed in practice, estimators based on subsampling present some drawbacks, as discussed in the introduction. In the following, we will introduce and analyse a novel approach for the drift estimation.

\begin{remark} Let us remark that for enhancing the clarity of the exposition, in this article we chose to focus on the case of a multi-dimensional parameter in the setting of one-dimensional diffusion processes. In fact, all the theory we present in the following could be generalized to the case of $d$-dimensional SDEs equivalent to \eqref{eq:SDE_MS}, which can be written as
	\begin{equation}
		\d X_t^\epl = - \sum_{i=1}^N \alpha_i \nabla V_i(X_t^\epl) \dd t - \frac1\epl \nabla p\left(\frac{X_t^\epl}{\epl}\right) \dd t + \sqrt{2\sigma} \dd W_t,
	\end{equation}
	where $W_t$ is a standard $d$-dimensional Brownian motion. The proof of all results below should be slightly modified, but we verified that all arguments still hold true in the $d$-dimensional case. 
\end{remark}

\section{The filtering approach}\label{sec:Filter}

In this section, we introduce and analyse a novel filtering approach to solve the biasedness issue highlighted by Theorem \ref{thm:Bias}. Let $\beta, \delta > 0$ and let us consider a family of exponential kernel functions $k \colon \R^+ \to \R$ defined as
\begin{equation}\label{eq:filter}
k(r) = C_\beta \delta^{-1/\beta} e^{-r^\beta/\delta},
\end{equation}
where $C_{\beta}$ is a normalizing constant given by
\begin{equation}
C_\beta = \beta \, \Gamma(1/\beta)^{-1},
\end{equation}
and where $\Gamma(\cdot)$ is the gamma function. We consider the process $Z^\epl \defeq (Z^\epl_t, 0 \leq t \leq T)$ defined by the weighted average
\begin{equation}\label{eq:ZDef}
	Z^{\epl}_t \defeq \int_0^t k(t - s)X^\epl_s \dd s.
\end{equation}
The process $Z^\epl$ can be interpreted as a smoothed version of the original trajectory $X^\epl$. In fact, in the field of signal processing the kernel \eqref{eq:filter} belongs to the class of low-pass linear time-invariant filters, which cut the high frequencies in a signal to highlight its slowest components. In the following, only in case $\beta = 1$ a rigorous analysis is carried on. Nonetheless, numerical experiments show that for higher values of $\beta$ the performances of estimators computed employing the filter are more robust and qualitatively better.

\begin{remark} Given a trajectory $X^\epl$, it is relatively inexpensive to compute $Z^\epl$ from a computational standpoint. In particular, the process $Z^\epl$ is the truncated convolution of the kernel with the process $X^\epl$. Hence, computational tools based on the Fast Fourier Transform (FFT) exist and allow to compute $Z^\epl$ fast component-wise. Moreover, the process $Z^\epl$ can be computed, in case $\beta = 1$, in a recursive manner and therefore ``online''.
\end{remark}

In the rest of this section we will focus on the properties of the process $Z^\epl$ when it is considered together with the original process $X^\epl$. In particular, we focus on ergodic properties and multiscale convergence. Finally, we present and analyse unbiased estimators obtained with the filtering process.

\subsection{Ergodic properties of the filter}\label{sec:ergodic}

Let us consider the filtering kernel \eqref{eq:filter} with $\beta = 1$, i.e.,
\begin{equation}\label{eq:filter_beta1}
	k(r) = \frac{1}{\delta} e^{-r/\delta}.
\end{equation}
In this case, Leibniz integral rule yields the equality
\begin{equation}
	\d Z^\epl_t = k(0) X^\epl_t \dd t + \int_0^t k'(t-s) X^\epl_s \dd s \dd t = \frac{1}{\delta} \left ( X^\epl_t - Z^\epl_t \right ) \dd t,
\end{equation}
which can be interpreted as an ordinary differential equation for $Z_t^\epl$ driven by the stochastic signal $X^\epl$. Considering the processes $X^\epl$ and $Z^\epl$ together, we obtain the system of two one-dimensional SDEs
\begin{equation}
\label{eq:systemSDE}
\begin{aligned}
\d X_t^\epl &= -\alpha \cdot V'(X_t^\epl) \dd t - \frac1\epl p'\left(\frac{X_t^\epl}\epl\right) \dd t+ \sqrt{2\sigma} \dd W_t, \\
\d Z^\epl_t &= \frac{1}{\delta} \left ( X^\epl_t - Z^\epl_t \right ) \dd t.
\end{aligned}
\end{equation}
The first ingredient for verifying the ergodic properties of the two-dimensional process $(X^\epl, Z^\epl)^\top \defeq ((X^\epl_t, Z^\epl_t)^\top, 0 \leq t \leq T)$ is verifying that the measure induced by the stochastic process admits a smooth density with respect to the Lebesgue measure. Since noise is present only on the first component, this is a consequence of the theory of hypo-ellipticity, as summarized in the following Lemma, \corr{whose proof is given in Appendix \ref{ap:Proofs}.}

\begin{lemma}\label{lem:density} Let $(X^\epl, Z^\epl)^\top$ be the solution of \eqref{eq:systemSDE} and let $\mu^\epl_t$ be the measure induced by the couple at time $t$. Then, the measure $\mu^\epl_t$ admits a smooth density $\rho^\epl_t$ with respect to the Lebesgue measure.
\end{lemma}

Once it is established that the law of the process admits a smooth density for all times $t>0$, which satisfies a time-dependent Fokker--Planck equation, we are interested in the limiting properties of this law. In particular, we know that the process $X^\epl$ alone is geometrically ergodic \cite[Theorem 4.4]{MSH02}, and we wish the couple $(X^\epl, Z^\epl)^\top$ to be endowed with the same property. The following Lemma guarantees that the couple is indeed geometrically ergodic, \corr{and its proof is given in Appendix \ref{ap:Proofs}.}

\begin{lemma}\label{lem:ergodicity} Let Assumption \ref{as:regularity} hold and let $b > 0$ be given in Assumption \ref{as:regularity}\ref{as:regularity_diss}. Then, if $\delta > 1/(4b)$, the process $(X^\epl, Z^\epl)^\top$ solution of \eqref{eq:systemSDE} is geometrically ergodic, i.e., there exists $C, \lambda > 0$ such that for all measurable $f\colon \R^2\to \R$ such that for some integer $q > 0$ 
	\begin{equation}
		f(x, z) \leq 1 + \norm{\begin{pmatrix} x & z \end{pmatrix}^\top}_2^q,
	\end{equation}
	it holds
	\begin{equation}
		\abs{\E f(X^\epl_t, Z^\epl_t) - \int_\R\int_\R f(x, z) \rho^\epl(x, z) \dd x \dd z} \leq C\left(1 + \norm{\begin{pmatrix} X^\epl_0 & Z^\epl_0 \end{pmatrix}^\top}_2^q \right)e^{-\lambda t},
	\end{equation}
	where $\E$ denotes expectation with respect to the Wiener measure and $\rho^\epl$ is the solution to the stationary Fokker--Planck equation
	\begin{equation}
	\label{eq:FPsystem}
	\sigma \partial^2_{xx} \rho^\epl(x,z) +  \partial_x \left( \left( \alpha \cdot V'(x) + \frac{1}{\epl} p' \left ( \frac{x}\epl \right ) \right ) \rho^\epl(x,z) \right) + \frac{1}{\delta} \partial_z \left((z - x) \rho^\epl(x,z)\right) = 0.
	\end{equation}
\end{lemma}

\begin{remark} The condition $\delta > 1 / (4b)$ is not very restrictive. Let the parameter dimension $N = 1$ and let $V(x) \propto x^{2p}$ for an integer $p > 1$. Then, Assumption \ref{as:regularity}\ref{as:regularity_diss} holds for an arbitrarily large $b > 0$. Therefore, the parameter of the filter $\delta$ can be chosen along the entire positive real axis. A similar argument can be employed for higher dimensions $N > 1$.
\end{remark}

\begin{example}\label{ex:OrnUhl} A closed form solution of \eqref{eq:FPsystem} can be obtained in a simple homogenized case with the dimension of the parameter $N=1$. Let $p(y) = 0$ and the parameters $\alpha$, $\sigma$ be replaced respectively by $A$ and $\Sigma$. Then, if $V(x) = x^2/2$, equation \eqref{eq:FPsystem} has the analytical solution 
\begin{equation}
\rho^0(x,z) = \frac{1}{C_{\rho^0}} \exp\left(-\frac{A}{\Sigma} \frac{x^2}{2} - \frac{1}{\delta \Sigma} \frac{(x - (1+A \delta)z)^2}{2}\right),
\end{equation}
where
\begin{equation}
C_{\rho^0} = \int_{\R} \int_{\R} \exp\left(-\frac{A}{\Sigma} \frac{x^2}{2} - \frac{1}{\delta \Sigma} \frac{(x - (1+A \delta)z)^2}{2}\right) \dd x \dd z.
\end{equation}
This is the density of a multivariate normal distribution $\mathcal N(0, \Gamma)$, where the covariance matrix is given by
\begin{equation}
\Gamma = \frac{\Sigma}{A (1 + A\delta)} \begin{pmatrix} 1+A\delta & 1 \\ 1 & 1 \end{pmatrix}.
\end{equation}
Let us remark that this distribution can be obtained from direct computations involving Gaussian processes. In particular, it is known that $X \sim \mathcal{GP}(\mu_t, \mathcal C(t, s))$, where at stationarity $\mu_t = 0$ and
\begin{equation}
	\mathcal C(t, s) = \frac{\Sigma}{A} e^{-A|t-s|}.
\end{equation}
The basic properties of Gaussian processes imply that $Z$ is a Gaussian process, and that the couple $(X, Z)^\top$ is a Gaussian process, too, whose mean and covariance are computable explicitly.
\end{example}

In a general case, it is not possible to find an explicit solution to \eqref{eq:FPsystem}. Nevertheless, it is possible to show some relevant properties of the solution itself, which are summarized in the following Lemma, \corr{whose proof is given in Appendix \ref{ap:Proofs}.}

\begin{lemma}\label{lem:FPMarginal} Under the assumptions of Lemma \ref{lem:ergodicity}, let $\rho^\epl$ be the solution of \eqref{eq:FPsystem} and let us write 
\begin{equation}\label{eq:densityDecomposition}
	\rho^\epl(x, z) = \phi^\epl(x)\psi^\epl(z)R^\epl(x,z),
\end{equation}
where $\phi^\epl$ and $\psi^\epl$ are the marginal densities of $X^\epl$ and $Z^\epl$ respectively, i.e., 
\begin{equation}
	\phi^\epl(x) = \int_{\R} \rho^\epl(x,z) \dd z, \quad  \psi^\epl(z) = \int_{\R} \rho^\epl(x,z) \dd x.
\end{equation}
Then, it holds
\begin{equation}\label{eq:marginalX}
	\phi^\epl(x) = \frac{1}{C_{\phi^\epl}} \exp\left(-\frac{1}{\sigma} \alpha \cdot V(x) - \frac{1}{\sigma} p \left ( \frac{x}\epl \right )\right),
\end{equation}
where
\begin{equation}
	C_{\phi^\epl} = \int_{\R} \exp\left(-\frac{1}{\sigma} \alpha \cdot V(x) - \frac{1}{\sigma} p \left ( \frac{x}\epl \right )\right) \dd x.
\end{equation}
Moreover, it holds
\begin{equation}
	\sigma \delta \int_{\R} \int_{\R} V'(z) \phi^\epl(x) \psi^\epl(z) \partial_x R^\epl(x,z) \dd x \dd z = \E^{\rho^\epl}[(X^\epl - Z^\epl)^2 V''(Z^\epl)].
\end{equation}
\end{lemma}

\subsection{Multiscale convergence}\label{sec:convMS}

We now investigate the convergence of the couple $(X^\epl, Z^\epl)^\top$ with respect to the multiscale parameter $\epl \to 0$. In particular, it is known that the invariant measure of $X^\epl$ converges weakly to the invariant measure of $X$ solution of the homogenized equation \eqref{eq:SDE_HOM}. The following result, \corr{whose proof is given in Appendix \ref{ap:Proofs},} guarantees the same kind of convergence for the couple $(X^\epl, Z^\epl)^\top$.

\begin{lemma}\label{lem:convMeasure} Under Assumption \ref{as:regularity}, let $\mu^\epl$ be the invariant measure of the couple $(X^\epl, Z^\epl)^\top$ and let $(X_0^\epl, Z_0^\epl)^\top \sim \mu^\epl$. Then, the measure $\mu^\epl$ converges weakly to the measure $\mu^0(\d x, \d z) = \rho^0(x, z) \dd x \dd z$, whose density $\rho^0$ is the unique solution of the Fokker--Planck equation
\begin{equation} \label{eq:FPsystem_homogenized}
	\Sigma \partial^2_{xx} \rho^0(x,z) + \partial_x\left( A \cdot V'(x) \rho^0(x,z) \right) + \frac{1}{\delta}\partial_z\left((z - x) \rho^0(x,z) \right) = 0,
\end{equation}
where $A$ and $\Sigma$ are the coefficients of the homogenized equation \eqref{eq:SDE_HOM}.
\end{lemma}

We conclude this section presenting an analogous result to Lemma \ref{lem:FPMarginal} for the limit distribution.

\begin{corollary}\label{lem:FPMarginal_Hom} Let $\rho^0$ be the solution of \eqref{eq:FPsystem_homogenized} and let us write 
	\begin{equation}
		\rho^0(x, z) = \phi^0(x)\psi^0(z)R^0(x,z),
	\end{equation}
	where $\phi^0$ and $\psi^0$ are the marginal densities, i.e., 
	\begin{equation}
		\phi^0(x) = \int_{\R} \rho^0(x,z) \dd z, \quad \psi^0(z) = \int_{\R} \rho^0(x,z) \dd x.
	\end{equation}
	Then, if $A$ and $\Sigma$ are the coefficients of the homogenized equation \eqref{eq:SDE_HOM}, it holds
	\begin{equation}
		\phi^0(x) = \frac{1}{C_{\phi^0}} \exp\left(- \frac1{\Sigma} A\cdot V(x)\right), \quad \text{where } \quad C_{\phi^0} = \int_{\R} \exp\left(- \frac1{\Sigma} A\cdot V(x) \right) \dd x.
	\end{equation}
	Moreover, it holds
	\begin{equation}
		\Sigma \delta \int_{\R} \int_{\R} V'(z) \phi^0(x) \psi^0(z) \partial_x R^0(x,z) \dd x \dd z = \E^{\rho^0}[(X - Z)^2 V''(Z)].
	\end{equation}
\end{corollary}
\begin{proof} The proof is directly obtained from Lemma \ref{lem:FPMarginal} replacing $p(y)=0$ and $\alpha, \sigma$ by $A, \Sigma$ respectively. 
\end{proof}

\subsection{The filtering-based estimator}\label{sec:FilterMLE}

We now consider the inference problem and propose our filtering-based estimator of the drift, which is formally given by the formula
\begin{equation}\label{eq:AHatMixed}
	\widehat A^\epl_k(T) = - \widetilde M^{-1} \widetilde h,
\end{equation}
where we employ the subscript $k$ for reference to the filter's kernel in \eqref{eq:filter}, and where
\begin{equation}\label{eq:MandHTilde}
	\widetilde M = \frac{1}{T} \int_0^T V'(Z^\epl_t) \otimes V'(X^\epl_t) \dd t, \qquad \text{and} \qquad \widetilde h = \frac{1}{T} \int_0^T V'(Z^\epl_t) \dd X^\epl_t.
\end{equation}
Let us remark that the formula above is obtained from \eqref{eq:MLE} by replacing only one instance of $X_t^\epl$ with $Z_t^\epl$ in both $M$ and $h$. In particular, it is fundamental for proving unbiasedness to keep in the definition of $h$ the differential of the original process $\d X^\epl_t$. Let us furthermore remark that $\widehat A^\epl_k(T)$ need not be the minimizer of some filtering-based likelihood function. In fact, if one were to replace $Z_t^\epl$ directly in \eqref{eq:Likelihood}, the symmetric part of the matrix $\widetilde M$ would appear and $\widehat A^\epl_k(T)$ would not be the minimizer. Therefore, the estimator $\widehat A^\epl_k(T)$ has to be thought of as a perturbation of $\widehat A^\epl(T)$, directly at the level of estimators and after the maximization procedure. The only theoretical guarantee which is still needed for the well-posedness of $\widehat A^\epl_k(T)$ is for $\widetilde M$ to be invertible, which we assume to be true and which we observed to hold in practice.

%\begin{lemma}\label{lem:AHatWellPosed} Under Assumption \ref{as:regularity}, the matrix $\widetilde M$ defined in \eqref{eq:MandHTilde} is invertible a.s. 
%\end{lemma}
%\begin{proof} 
%\todo: \corr{Idea. We don't need $\widetilde M$ to be symmetric positive definite (it is not) due to the comment just above the lemma. Nevertheless, if the functions $V_i'(x)$ are not such that the rows of $\widetilde M$ are identical for any $(X_t, Z_t)$, then the randomness in the entries should guarantee that $\det \widetilde M \neq 0$ with probability one. Still missing the (probably trivial) argument why this is true.}
%\end{proof}

We are now able to present the main result of unbiasedness of the filtering-based estimator.
\begin{theorem}\label{thm:mainTheorem} Let the assumptions of Lemma \ref{lem:ergodicity} and Lemma \ref{lem:convMeasure} hold, and let $\widehat A^\epl_k(T)$ be defined in \eqref{eq:AHatMixed}. If $\widetilde M$ is invertible, then
	\begin{equation}
	\lim_{\epl \to 0} \lim_{T \to \infty} \widehat A_k^\epl(T) = A, \quad \text{a.s.},
	\end{equation}
	where $A$ is the drift coefficient of the homogenized equation \eqref{eq:SDE_MS}.
\end{theorem}

\begin{proof} Replacing the expression of $\d X^\epl_t$ into \eqref{eq:MandHTilde}, we get for $\widetilde h$
\begin{equation}\label{Ahat_decomposition}
\widetilde h = -\widetilde M \alpha - \frac{1}{T} \int_0^T \frac{1}{\epl} p' \left ( \frac{X^\epl_t}{\epl} \right ) V'(Z^\epl_t) \dd t + \frac{\sqrt{2\sigma}}{T} \int_0^T V'(Z^\epl_t) \dd W_t.
Therefore, we have
\begin{equation}\label{eq:alphaDecomposition}
	\widehat A^\epl_k(T) &= \alpha + \frac{1}{T} \widetilde M^{-1} \int_0^T \frac{1}{\epl} p' \left ( \frac{X^\epl_t}{\epl} \right ) V'(Z^\epl_t) \dd t - \frac{\sqrt{2\sigma}}{T}  \widetilde M^{-1} \int_0^T V'(Z^\epl_t) \dd W_t\\
	&\eqdef \alpha + I_1^\epl(T) - I_2^\epl(T).
We study the terms $I_1^\epl(T)$ and $I_2^\epl(T)$ separately. The ergodic theorem applied to $I_1^\epl(T)$ yields
\begin{equation}\label{limit_estimator}
\lim_{T \to \infty} I_1^\epl(T) = \E^{\rho^\epl} [V'(Z^\epl) \otimes V'(X^\epl)]^{-1} \E^{\rho^\epl} \left [ \frac{1}{\epl} p' \left ( \frac{X^\epl}{\epl} \right ) V'(Z^\epl) \right ], \quad \text{a.s.}
Due to Lemma \ref{lem:FPMarginal} and integrating by parts, we have
	\E^{\rho^\epl} \left [ \frac{1}{\epl} p' \left ( \frac{X^\epl}{\epl} \right ) V'(Z^\epl) \right ] &= \int_{\R} \int_{\R} V'(z) \, \frac1\epl p' \left ( \frac{x}\epl \right ) \frac{1}{C_{\phi^\epl}} e^{- \frac{1}{\sigma} \alpha \cdot V(x)} e^{ - \frac{1}{\sigma} p \left ( \frac{x}\epl \right)} \psi^\epl(z) R^\epl(x,z) \dd x \dd z \\

>>>>

\input{../ex_shared}

\begin{document}
